---
title: "US Mass Shootings Data Analysis"
output:
  html_document: default
  html_notebook: default
---

## 1. Loading The Dataset
```{r message=FALSE, warning=FALSE}
library(ggplot2)
library(scales)
library(tidyr)
library(dplyr)
library(formattable)
setwd("E:/MOOC")
shootings <- read.csv("MassShootingsDataset.csv",stringsAsFactors=FALSE)
```

## 2. Dimensions Of The Dataset
```{r}
dim(shootings)
```

## 3. Description of the variables and thier types.
```{r}
str(shootings)
```

## 4. Overall Metrics
```{r}
total_casualties <- sum(shootings$Fatalities)
total_injured <- sum(shootings$Injured)
total_victims <- sum(shootings$Total.victims)

sprintf("Total Number Of Casualties Reported Till Date: %s",total_casualties)
sprintf("Total Number Of Injured Reported Till Date: %s",total_injured)
sprintf("Total Number Of Victims Reported Till Date: %s",total_victims)
```

## 5. Top 10 Mass Shootings
```{r}
shootings$Date <- as.Date(shootings$Date,"%m/%d/%Y")
shootings$Year <- format(shootings$Date,"%Y")
newData <- head(shootings[order(-shootings$Total.victims),],10)
row.names(newData) <- NULL
formattable(subset(newData, select = c(Title,Location,Total.victims,Year)),row.names = FALSE)
```

## 6. Timeline Analysis

```{r fig.width = 11, fig.height = 5}

ggplot(shootings,aes(x=Year))+
  geom_bar(fill="#A52A2A")+
  geom_text(stat='count',aes(label=..count..),vjust=-1)+
  scale_y_continuous(breaks = seq(0,80,10),limits = c(0,80))+
  labs(x="Year",y="Number Of Attacks",title="Number Of Attacks Per Year")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  theme(plot.title = element_text(hjust = 0.5,size = 14))
```
 * The above timeline analysis shows that highest number of attacks were carried out in the year 2015 and 2016, Now lets compare total.victims over a year timeline to see whether year 2015 and 2016 report high casualties too. 
```{r fig.width = 10, fig.height = 5}
ggplot(shootings,aes(x=Year,group=1))+
  geom_line(stat="summary",fun.y = sum,aes(y=Total.victims),color= "red")+
  geom_point(stat="summary",fun.y = sum,aes(y=Total.victims))+
  scale_y_continuous(breaks = seq(0,700,50))+
  labs(x="Year",y="Number Of Victims",title="Timeline Comparison of Victims Per Year")+
  theme_bw()+
  theme(axis.text.x = element_text(hjust = 1,angle = 60))+
  theme(plot.title = element_text(hjust = 0.5,size = 14))
```

 * As expected, very high number of victims were reported in year 2015 and 2016, but 2017 inspite of having very less attacks number of attacks, reported the highest number of victims because of "Las Vegas Strip mass shooting" in which around 526 people were injured and 58 people were killed. Now lets compare the total casualties over a year timeline.
 
```{r fig.width = 10, fig.height = 5}
ggplot(shootings,aes(x=Year,group=1))+
  geom_line(stat="summary",fun.y=sum,aes(y=Fatalities),color="red",size=0.5)+
  geom_point(stat="summary",fun.y=sum,aes(y=Fatalities))+
  labs(x="Year",y="Number Of Casualties Reported",title="Timeline Comparison Of Casualties Per Year")+
  scale_y_continuous(breaks = seq(0,350,50),limits = c(0,250))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 60,hjust = 1))+
  theme(plot.title = element_text(size = 14,hjust = 0.5))
```

```{r fig.width = 10, fig.height = 5}
ggplot(shootings,aes(x=Year,group=1))+
  geom_line(stat="summary",fun.y=sum,aes(y=Injured),color="red",size=0.5)+
  geom_point(stat="summary",fun.y=sum,aes(y=Injured))+
  labs(x="Year",y="Number Of Injured Reported",title="Timeline Comparison Of Injured Per Year")+
  scale_y_continuous(breaks = seq(0,550,50),limits = c(0,550))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 60,hjust = 1))+
  theme(plot.title = element_text(size = 14,hjust = 0.5))
```

Lets combine the above metrics into a single graph which represents both the Fatalties and injured per year.
```{r fig.width = 12, fig.height = 7}
long_data <- gather(subset(shootings,select = c(Year,Fatalities,Injured)),VictimType,NoOfVictims,Fatalities,Injured)

ggplot(long_data,aes(x=Year))+
  geom_bar(stat = "summary",fun.y=sum,aes(y=NoOfVictims,fill=VictimType),width = 0.6)+
  scale_y_continuous(breaks = seq(0,650,50),limits = c(0,650))+
  labs(x="Year",y="Number Of Attacks",title="Timeline Comparison Of Victims Per Year")+
  theme_bw()+
  #scale_fill_manual(values = c("red","blue"))+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  theme(plot.title = element_text(hjust = 0.5,size = 14))
```

## 7. Mental Health Issues Exploration.
 * Lets take a look at the data to identify anamolies. 
```{r}
table(shootings$Mental.Health.Issues)
```

 * The data seams to contain duplicates for "Unknown" value and as "Unknown" and "Unclear" mean the same, lets rename the column values such that they contain consistent values.
```{r}
shootings$Mental.Health.Issues <- ifelse(shootings$Mental.Health.Issues == "unknown" | shootings$Mental.Health.Issues == "Unclear","Unknown",shootings$Mental.Health.Issues)
table(shootings$Mental.Health.Issues)
```
```{r}
ggplot(shootings,aes(x=Mental.Health.Issues))+
  geom_bar(width = 0.1,fill="#A52A2A")+
  geom_text(stat='count',aes(label=..count..),vjust=-1)+
  scale_y_continuous(limits = c(0,130),breaks = seq(0,130,20))+
  labs(x="Has Mental Health Issues",y="No Of People",title = "Mental Health Issues Comparison")+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5,size = 14))
```

## 8. Gender Based Exploration.
```{r}
table(shootings$Gender)
```
 * The Gender data seems to contain few duplicates, Lets elminate these duplicates, so that data represents consistent values.
```{r}
shootings$Gender <- ifelse(shootings$Gender == "M","Male",ifelse(shootings$Gender == "M/F","Male/Female",shootings$Gender))
table(shootings$Gender)
```
```{r}
ggplot(shootings,aes(x=Gender))+
  geom_bar(width = 0.2,fill="#A52A2A")+
  geom_text(stat='count',aes(label=..count..),hjust=-0.7)+
  scale_y_continuous(limits = c(0,310),breaks = seq(0,310,40))+
  labs(x="Gender Type",y="No Of Attacks Carried Out",title = "Gender Based Comparison Of Attacks")+
  coord_flip()+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5,size = 14))
```
 
 * The Gender based exploration shows that most of the attacks were carried out by males. Now lets compare no of fatalties caused by gender.
```{r}
ggplot(shootings,aes(x=Gender,y=Total.victims))+
  geom_bar(stat="summary",fun.y=sum,fill="#A52A2A",width=0.2)+
  labs(x="Gender",y="Number Of Victims",title = "Comparison Of victims By Attacker's Gender Type")+
  theme_bw()+
  coord_flip()+
  theme(plot.title = element_text(hjust = 0.5,size = 14))
```

## 9. Province Based Exploration  
```{r}
shootings$Province <- ""
value <- ","
for(i in 1:nrow(shootings)){
 shootings[i,15] <-ifelse(shootings[i,3] != "",strsplit(shootings[i,3],",")[[1]][2],"")  
}
```

```{r fig.width = 8, fig.height = 10}
shootings$Province <- ordered(shootings$Province,levels = names(sort(table(shootings$Province))))
ggplot(subset(shootings,Province != ""),aes(Province))+
  geom_bar(width = 0.4,fill="#A52A2A")+
  geom_text(stat='count',aes(label=..count..),hjust=-0.7)+
  scale_y_continuous(limits = c(0,30),breaks = seq(0,30,10))+
  labs(x="Province",y="Number Of Shootings",title = "Province Based Comparison Of Shootings")+
  coord_flip()+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5,size = 14))
```

  * The Provice based comparison shows that most number of attacks were carried out in calfornia state. Now lets compare the provinces based number of victims. 

```{r fig.width = 8, fig.height = 10}
victims_by_province <- subset(shootings,Province != "") %>% 
  group_by(Province) %>%
  summarise(total=sum(Total.victims),n = n())%>%
  arrange(desc(total))

ggplot(victims_by_province,aes(x=reorder(Province,total),y=total))+
  geom_bar(stat="identity",width = 0.4,fill="#A52A2A")+
  geom_text(aes(label=total,hjust=-0.4))+
  labs(x="Province",y="Number Of Victims",title = "Province Based Comparison Of Victims")+
  scale_y_continuous(limits = c(0,600))+
  coord_flip()+
  theme_bw()+
  theme(plot.title = element_text(hjust = 0.5,size = 14))
```

 * The provine based comparison of victims show that most effected state is Naveda. Naveda reported high number of victims because it recenctly witnesed a mass shooting which was one of the most horrible attack in USA's history. If we exclude that attack then most number of victims and attacks were reported by the state of California.     
 
## 10. Race Based Exploration
```{r}
table(shootings$Race)
```
 
 * The Race data column seems to have some duplicate values.Lets eliminate these duplicate values, so that column represents consistent values.
 
```{r}
shootings$Race <- ifelse(shootings$Race == "white","White",ifelse(shootings$Race == "black","Black",ifelse(shootings$Race == "" | shootings$Race == "Some other race" | shootings$Race == "Unknown", "Other",shootings$Race)))

table(shootings$Race)
```


```{r fig.width = 10, fig.height = 5}
ggplot(shootings,aes(x=Race))+
  geom_bar(fill="#A52A2A",width = 0.5)+
  labs(x="Race",y="Number Of Attacks Carried Out",title = "Race Based Comparison Of Attacks")+
  theme_bw()+
  coord_flip()+
  theme(plot.title = element_text(hjust = 0.5,size = 14))
```

```{r fig.width = 10, fig.height = 5}
ggplot(shootings,aes(x=Race,y=Total.victims))+
  geom_bar(stat="summary",fun.y=sum,fill="#A52A2A",width=0.5)+
  labs(x="Race",y="Number Of Victims",title = "Comparison Of victims Based On Attacker's Race Type")+
  theme_bw()+
  coord_flip()+
  theme(plot.title = element_text(hjust = 0.5,size = 14))
```

## 11. Gender Based Analysis Of Shootings Over Year Timeline
```{r fig.width = 12, fig.height = 5}
ggplot(shootings,aes(x=Year))+
  geom_bar(aes(fill=Gender))+
  geom_text(stat='count',aes(label=..count..),vjust=-1)+
  scale_y_continuous(breaks = seq(0,80,10),limits = c(0,80))+
  labs(x="Year",y="Number Of Attacks",title="Gender Based Analysis Of Shootings Over Year Timeline")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 60, hjust = 1))+
  theme(plot.title = element_text(hjust = 0.5,size = 14))
```

 * The Gender based analysis of shootings shows that most of the attacks were carried out by the males every year.
 
 
 
 
 
